---
title: "Homework 2"
author: "Painter, Ty"
date: "`r date()`"
output: github_document
---
# Using the RMarkdown/knitr/github mechanism, implement the following tasks:
- Use the prostate cancer data.
```{r}
prostate <-read.table(url('https://web.stanford.edu/~hastie/ElemStatLearn/datasets/prostate.data'))
prostate <- subset(prostate, train==TRUE) # 2/3 for train, 1/3 test
```

- Use the cor function to reproduce the correlations listed in HTF Table 3.1, page 50.
```{r}
prostate.x <- prostate[,c("lweight", "age", "lbph", "svi", "lcp", "gleason", "pgg45")]
prostate.y <- prostate[,c("lcavol", "lweight", "age", "lbph", "svi", "lcp", "gleason")]
cor_table <- round(cor(prostate.x, prostate.y),3)

cor_table[upper.tri(cor_table)]<-""
cor_table<-as.data.frame(cor_table)
cor_table
```

- Treat lcavol as the outcome, and use all other variables in the data set as predictors.
- With the training subset of the prostate data, train a least-squares regression model with all predictors using the lm function.
```{r}

```

- Use the testing subset to compute the test error (average squared-error loss) using the fitted least-squares regression model.
- Train a ridge regression model using the glmnet function, and tune the value of lambda (i.e., use guess and check to find the value of lambda that approximately minimizes the test error).
- Create a figure that shows the training and test error associated with ridge regression as a function of lambda
- Create a path diagram of the ridge regression analysis, similar to HTF Figure 3.8
## Links
(https://lagunita.stanford.edu/courses/HumanitiesSciences/StatLearning/Winter2016/about)