---
title: "Homework 2"
author: "Painter, Ty"
date: "`r date()`"
output: github_document
---

**Using the RMarkdown/knitr/github mechanism, implement the following tasks by extending the example R script ( prostate-data-lin.R):** 

```{r}
library(devtools)
## load prostate data

prostate <-read.table(url('https://web.stanford.edu/~hastie/ElemStatLearn/datasets/prostate.data'))

## subset to training examples
prostate_train <- subset(prostate, train==TRUE) # 2/3 for train, 1/3 test

## plot lcavol vs lpsa
plot_psa_data <- function(dat=prostate_train) {
  plot(dat$lpsa, dat$lcavol,
       xlab="log Prostate Screening Antigen (lpsa)",
       ylab="log Cancer Volume (lcavol)",
       pch=16)
}
plot_psa_data()
```

- Write functions that implement the L1 loss and tilted absolute loss functions.

```{r}
## L2 loss function, squared error loss
L2_loss <- function(y, yhat){
  (y-yhat)^2
}

## L1 loss function, absolute error loss
L1_loss <- function(y, yhat){
  abs(y-yhat)
}

## tilted absolute loss function
tilt_loss <- function(y, yhat, tau){
 tav = ifelse(y-yhat>0, tau*(y-yhat), (tau-1)*(y-yhat)) # gives absolute loss
 return(tav)
}


## fit simple linear model using numerical optimization
fit_lin <- function(y, x, loss, beta_init = c(-0.51, 0.75),...) { # estimates beta1 and beta2; beta minimizes residual squares
  err <- function(beta) # calculates avg L2 loss function
    mean(loss(y,  beta[1] + beta[2]*x, ...)) # yhat = beta[1] + beta[2]*x
  
  beta <- optim(par = beta_init, fn = err) # want to minimize fn, optimize between (-0.51, 0.75), err is initial value 
  
  return(beta) # values of beta that minimize training error, uses gradient method
}

## make predictions from linear model
predict_lin <- function(x, beta) {
  beta[1] + beta[2]*x
}

## fit linear model - estimates betas to minimize training error
lin_beta_L2 <- fit_lin(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,
                    loss=L2_loss)

## fit linear model - estimates betas to minimize training error
lin_beta_L1 <- fit_lin(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,
                    loss=L1_loss)

## fit linear model - estimates betas to minimize training error
lin_beta_tilt25 <- fit_lin(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,
                    loss=tilt_loss,
                    tau=0.25) 

## fit linear model - estimates betas to minimize training error
lin_beta_tilt75 <- fit_lin(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,
                    loss=tilt_loss,
                    tau=0.75)

## compute predictions for a grid of inputs
x_grid <- seq(min(prostate_train$lpsa),
              max(prostate_train$lpsa),
              length.out=100)
y_grid <- seq(from=min(prostate_train$lcavol),
              to=max(prostate_train$lcavol),
              length.out=100)

lin_pred_L2 <- predict_lin(x=x_grid, beta=lin_beta_L2$par) # generates yhat (yhat = beta[1] + beta[2]*x)
lin_pred_L1 <- predict_lin(x=x_grid, beta=lin_beta_L1$par) # generates yhat (yhat = beta[1] + beta[2]*x)
lin_pred_tilt25 <- predict_lin(x=x_grid, beta=lin_beta_tilt25$par) # generates yhat (yhat = beta[1] + beta[2]*x)
lin_pred_tilt75 <- predict_lin(x=x_grid, beta=lin_beta_tilt75$par) # generates yhat (yhat = beta[1] + beta[2]*x)
```

- Create a figure that shows lpsa (x-axis) versus lcavol (y-axis). Add and label (using the 'legend' function) the linear model predictors associated with L2 loss, L1 loss, and tilted absolute value loss for tau = 0.25 and 0.75.

```{r}
## plot data
plot_psa_data()
## plot L1, L2, tilted abs value predictions
lines(x=x_grid, y=lin_pred_L2, col="red", lwd=2)
lines(x=x_grid, y=lin_pred_L1, col="blue",lwd=2)
lines(x=x_grid, y=lin_pred_tilt25, col="green",lwd=2)
lines(x=x_grid, y=lin_pred_tilt75, col="orange",lwd=2)
legend("topleft",
       legend = c("L2 Loss", "L1 Loss", "Tilted Abs. Loss (tau=0.25)", "Tilted Abs. Loss (tau=0.75)"),
       col = c("red", "blue", "green", "orange"),
       lty = 1, lwd=2)
```

- Write functions to fit and predict from a simple nonlinear model with three parameters defined by 'beta[1] + beta[2]*exp(-beta[3]*x)'. Hint: make copies of 'fit_lin' and 'predict_lin' and modify them to fit the nonlinear model. Use c(-1.0, 0.0, -0.3) as 'beta_init'.

```{r}
## L2 loss function, squared error loss
L2_loss <- function(y, yhat)
  (y-yhat)^2

## L1 loss function, absolute error loss
L1_loss <- function(y, yhat)
  abs(y-yhat)

## tilted absolute loss function
tilt_loss <- function(y, yhat, tau){
 tav = ifelse(y-yhat>0, tau*(y-yhat), (tau-1)*(y-yhat)) # gives absolute loss
 return(tav)
}

## fit simple nonlinear model using numerical optimization
fit_nlin <- function(y, x, loss, beta_init = c(-1.0, 0.0, -0.3), ...) { # estimates beta1 and beta2; beta minimizes residual squares
  err <- function(beta) # calculates avg L2 loss function
    mean(loss(y,  beta[1] + beta[2]*exp(-beta[3]*x),...)) # yhat = beta[1] + beta[2]*exp(-beta[3]*x)
  
  beta <- optim(par = beta_init, fn = err) # want to minimize fn, optimize between (-1.0, 0.0, -0.3), err is initial value 
  
  return(beta) # values of beta that minimize training error, uses gradient method
}

## make predictions from nonlinear model
predict_lin_NL <- function(x, beta)
  beta[1] + beta[2]*exp(-beta[3]*x)

## fit nonlinear model - estimates betas to minimize training error
lin_beta_L2_NL <- fit_nlin(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,
                    loss=L2_loss)

## fit nonlinear model - estimates betas to minimize training error
lin_beta_L1_NL <- fit_nlin(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,
                    loss=L1_loss)

## fit nonlinear model - estimates betas to minimize training error
lin_beta_tilt25_NL <- fit_nlin(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,
                    loss=tilt_loss,
                    tau=0.25)

## fit nonlinear model - estimates betas to minimize training error
lin_beta_tilt75_NL <- fit_nlin(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,
                    loss=tilt_loss,
                    tau=0.75)

## compute predictions for a grid of inputs
x_grid <- seq(min(prostate_train$lpsa),
              max(prostate_train$lpsa),
              length.out=100)
y_grid <- seq(from=min(prostate_train$lcavol),
              to=max(prostate_train$lcavol),
              length.out=100)


lin_pred_L2_NL <- predict_lin_NL(x=x_grid, beta=lin_beta_L2_NL$par) # generates yhat (yhat = beta[1] + beta[2]*x)
lin_pred_L1_NL <- predict_lin_NL(x=x_grid, beta=lin_beta_L1_NL$par) # generates yhat (yhat = beta[1] + beta[2]*x)
lin_pred_tilt25_NL <- predict_lin_NL(x=x_grid, beta=lin_beta_tilt25_NL$par) # generates yhat (yhat = beta[1] + beta[2]*x)
lin_pred_tilt75_NL <- predict_lin_NL(x=x_grid, beta=lin_beta_tilt75_NL$par) # generates yhat (yhat = beta[1] + beta[2]*x)
```

- Create a figure that shows lpsa (x-axis) versus lcavol (y-axis). Add and label (using the 'legend' function) the nonlinear model predictors associated with L2 loss, L1 loss, and tilted absolute value loss for tau = 0.25 and 0.75.

```{r}
## plot data
plot_psa_data()
## plot L1, L2, tilted absolute predictions
lines(x=x_grid, y=lin_pred_L2_NL, col="red", lwd=2)
lines(x=x_grid, y=lin_pred_L1_NL, col="blue", lwd=2)
lines(x=x_grid, y=lin_pred_tilt25_NL, col="green",lwd=2)
lines(x=x_grid, y=lin_pred_tilt75_NL, col="orange",lwd=2)
legend("topleft",
       legend = c("L2 Loss", "L1 Loss", "Tilted Abs. Loss (tau=0.25)", "Tilted Abs. Loss (tau=0.75)"),
       col = c("red", "blue", "green", "orange"),
       lty = 1, lwd=2)
```